Multisequence algorithm for coarse-grained biomolecular simulations: application to protein fold switching

### 2017-05-22 Manuscript.tex
Methods

\newcommand	 {\sbar}	{{s}}
\newcommand	 {\rbar}	{{r}}
\newcommand	 {\hi}		{{h_\mathrm{i}}}
\newcommand	 {\pii}  	{{p_\mathrm{i}}}
\newcommand	 {\kb}		{{k_\mathrm{B}}}
\newcommand	 {\Tlow}	{{T_\mathrm{low}}}
\newcommand 	 {\Pnat} 	{{P_\mathrm{nat}}}
\newcommand {\Qa}		{{Q_\mathrm{A}}}
\newcommand {\Qb}		{{Q_\mathrm{B}}}
\newcommand {\Pcut}     	{{P_\mathrm{cut}}}
\newcommand {\TlowI}     {{T^\mathrm{I}_\mathrm{low}}}
\newcommand {\TlowII}    {{T^\mathrm{II}_\mathrm{low}}}
\newcommand {\Ptot}	{{P_\mathrm{tot}}}
\newcommand {\PaI}    	{{P_\mathrm{IA}}}
\newcommand {\PbI}    	{{P_\mathrm{IB}}}
\newcommand {\PaII}    	{{P_\mathrm{IIA}}}
\newcommand {\PbII}    	{{P_\mathrm{IIB}}}
\newcommand {\Eave}	{{E_\mathrm{ave}}}
\newcommand {\sigE}	{{\sigma_{\left < E \right >}}}
\newcommand {\SR}		{${\mathrm{S16}_{144}}$}
\newcommand {\SI}		{${\mathrm{S16}_{1024}}$}	
\newcommand {\SII}		{${\mathrm{S16}_{1024}}$}

@@

\subsection{Stability measures}
\noindent
To quantify the stabilities of various folds, we use the measures introduced in Ref.~\cite{Holzgrafe2015}. We classify conformations $r$ as part of a fold if $Q\ge0.8$ where $Q\in [0,1]$ quantifies the similarity of $r$ to that fold.

For folds A and B, $Q$ is determined by defining a set of backbone-backbone hydrogen bonds that are present in those folds (ignoring N and C terminal amino acids). For fold A, the fold-specific set of contacts (i,j) between the CO group of amino acid i and the NH group of amino acid j is $\{$(2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15)$\}$, and for fold B the contact set is $\{$(3,14), (5,12), (7,10), (10,7), (12,5), (14,3)$\}$. A hydrogen bond is considered formed if (i) the NHO and COH angles both are $>90^\circ$ and (ii) the HO distance is $<2.7$~{\AA}.

%\begin{equation}
%Q_\mathrm{2\alpha} = ( Q_\alpha^\mathrm{1-16} + Q_\alpha^\mathrm{20-35} + Q_\mathrm{tert}) ) / 3
%\end{equation}
%
%\begin{equation}
%Q_\mathrm{\alpha\beta} = ( Q_\alpha^\mathrm{1-16} + Q_\beta^\mathrm{20-35} + Q_\mathrm{tert} ) / 3
%\end{equation}

@@

\vspace{12pt}
\subsection{A1-N1 and A2-TN binary sequence spaces}
\noindent
We now apply the MS algorithm to the sets {\SI} and {\SII} which represent the full binary sequence spaces for the A1-N1 and A2-TN systems, respectively. Both sets include $2^{10}=1,024$ sequences. Using MS simulations, we determined the thermodynamic behavior at the temperatures $\TlowI$ and $\TlowII$ for the A1-N1 and A2-TN systems, respectively (see Table~1). In particular, this allow us to find the stabilities of folds IA and IB ($\PaI$ and $\PbI$) for all sequences in {\SI} and the stabilities of folds IIA and IIB ($\PaII$ and $\PbII$) for all sequences in {\SII}.

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to IA-IB or IIA-IIB fold changes without passing through an unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at only one amino acid position. To determine if a sequence is stable we used the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PaI + \PbI$ and $\PaII + \PbII$ for A1-N1 and A2-AN1, respectively. The precise network obtained depends, of course, on the cut-off value of $\Pcut$. In general, a higher $\Pcut$ means a selection of more stable pathways. Fig.~5 shows the networks obtained with $\Pcut=0.50$. The number of viable mutational pathways that connect the start and end points are 516,972 between A1 and N1 and  57,912 between A2 and AN1. Because there are $10! =3,628,800$ possible mutational pathways between start and end points in both cases, these pathways represent 14.2 \% and 1.6 \% of all possible pathways in A1-N1 and A2-AN1, respectively. Hence, A1 and N1 are rather highly connected at this stability threshold. For $\Pcut=0.60$, the numbers are 104,640 pathways (2.9\%) for A1-N1 and 22,512 (0.6\%) pathways for A2-AN1. We find that there are no possible A1-to-N1 or A2-to-AN1 pathways when $\Pcut\ge0.74$ and $\Pcut\ge 0.66$, respectively. \\

@@

\section{Discussion}
\noindent
We have evaluated a generalized-ensemble algorithm for coarse-grained simulations of biomolecules that works by making the sequence a dynamic parameter and thereby leading to a random walk in sequence space. We find that this approach has two main benefits. On the one hand, it provides a convenient way to sample the respective canonical distributions for a set of sequences in a single run and, on the other, it also enhances the sampling of conformational space.

The sampling efficiency of the MS algorithm can be assessed by the comparison with ST in Figure~3. At the highest studied temperatures, which are well above the folding temperatures of both A1 and N1, the statistical errors on the average total energy, $\sigE$, are very similar for the two methods. This is not unexpected because conformational sampling of short polymers at high $T$ does not involve crossing any major energy barriers. As a result, successively sampled conformations, for a given combination of sequence $\sbar$ and temperature $T$, are therefore likely mostly uncorrelated in both methods which in turn leads to similar $\sigE$ values.

At lower temperatures, we find that the MS simulations often yields significantly smaller $\sigE$ than ST simulations. It is notable that this acceleration in conformational sampling is achieved despite that simulations are carried out at constant temperature. Hence, rather than promoting escape from local minima in the free energy landscape by visits to higher temperature, as in ST~\cite{Marinari1992,Lyubartsev1992} or replica-exchange,~\cite{Swendsen1986} the MS algorithm promotes local minima escape through visits to other sequences for which the minima may not be as severe or are absent altogether. For the same reason, the efficiency of the MS algorithm likely depends on the size and character of sequences included in the simulation. In particular, the MS simulation is likely dependent on the inclusion of at least some less stable sequences to promote at least partial unfolding of the chain and the transition to different conformational states.

Clearly, the MS method should not be seen as a general simulation method to speed up conformational sampling due to the complexity of performing sequence updates in most biomolecular models. However, our results indicate that for coarse-grained models that permit the sequence updates as a Metropolis step and when visits to higher $T$ is unwanted (or unnecessary), the MS method can be highly efficient. A notable feature of the algorithm is that it provides escape from local minima on the free energy landscape of proteins without resorting to a change in $T$. This feature could make it attractive as an alternative to exploring the equilibrium behavior of proteins and peptides in the context or ordered dense phases such as lipid bilayers, where escape from local minima are especially challenging even in CG simulations.~\cite{Bereau2015} Simulations of lipid bilayers at elevated $T$ must typically be avoided because it can lead to unwanted perturbations of the membrane structure, although it can be a viable approach to speed up conformational sampling in particular cases.~\cite{Ulmschneider2010}

>>>>

\newcommand	 {\sbar}	{{s}}
\newcommand	 {\rbar}	{{r}}
\newcommand	 {\hi}		{{h_\mathrm{i}}}
\newcommand	 {\pii}  	{{p_\mathrm{i}}}
\newcommand	 {\kb}		{{k_\mathrm{B}}}
\newcommand	 {\Tlow}	{{T_\mathrm{low}}}
\newcommand 	 {\Pnat} 	{{P_\mathrm{nat}}}
\newcommand {\QIA}	{{Q_\mathrm{IA}}}
\newcommand {\QIB}	{{Q_\mathrm{IB}}}
\newcommand {\QIIA}	{{Q_\mathrm{IIA}}}
\newcommand {\QIIB}	{{Q_\mathrm{IIB}}}
\newcommand {\Pcut}     	{{P_\mathrm{cut}}}
\newcommand {\TlowI}     {{T^\mathrm{I}_\mathrm{low}}}
\newcommand {\TlowII}    {{T^\mathrm{II}_\mathrm{low}}}
\newcommand {\Ptot}	{{P_\mathrm{tot}}}
\newcommand {\PIA}    	{{P_\mathrm{IA}}}
\newcommand {\PIB}    	{{P_\mathrm{IB}}}
\newcommand {\PIIA}    	{{P_\mathrm{IIA}}}
\newcommand {\PIIB}    	{{P_\mathrm{IIB}}}
\newcommand {\Eave}	{{E_\mathrm{ave}}}
\newcommand {\sigE}	{{\sigma_{\left < E \right >}}}
\newcommand {\SR}		{${\mathrm{S16}_{144}}$}
\newcommand {\SI}		{${\mathrm{S16}_{1024}}$}	
\newcommand {\SII}		{${\mathrm{S16}_{1024}}$}

@@

\subsection{Observables}
\noindent
Fold stabilities are calculated as in Ref.~\citenum{Holzgrafe2015} and also described briefly below. We define first two structural similarity measures $\QIA$ and $\QIB$ for folds IA and IB, respectively, indicating what fraction of the fold-specific backbone-backbone hydrogen bonds have been formed. The fold IA-hydrogen bonds are (2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15) and the fold IB-bonds are (3,14), (5,12), (7,10), (10,7), (12,5), (14,3), where (i,j) indicates a hydrogen bond between the CO group of amino acid i and the NH group of amino acid j. The  stabilities of fold IA and IB are then defined as the probabilities $\PIA = P(\QIA\ge0.8)$ and $\PIB = P(\QIB\ge0.80)$, respectively, i.e., stability is the probability that at least 80\% of the fold's hydrogen bonds are formed. $\PIA$ and $\PIB$ depend on both sequence $\sbar$ and temperature $T$. For example, $\PIA=0.xx$ for A1 and $\PIB=0.xx$ for N1 at $\kb T = 0.43$. Structural similarity measures for 35-amino acid folds IIA and IIB are defined as $\QIIA = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IA}^\mathrm{20-35} + Q_\mathrm{tert}) ) / 3$ and $\QIIB = ( Q_\mathrm{IA}^\mathrm{1-16} + Q_\mathrm{IB}^\mathrm{20-35} + Q_\mathrm{tert} ) / 3$, respectively, where superscripts on $\QIA$ and $\QIB$ indicate over which amino acid positions those measures are applied to within the longer 35 amino acid sequences and $Q_\mathrm{tert}$ is a measure that counts the number of $C_\beta$-$C_\beta$ contacts between the two secondary structure elements of these folds.~\cite{Holzgrafe2015} As for IA and IB, the stabilities of folds IIA and IIB are defined as $\PIIA = P(\QIIA\ge0.8)$ and $\PIIB = P(\QIIB\ge0.80)$, respectively. 
%In calculating $Q$ values, a hydrogen bonßd is considered formed if (i) the NHO and COH angles are both $>90^\circ$ and (ii) the HO distance is $<2.7$~{\AA}.

@@

\vspace{12pt}
\subsection{A1-N1 and A2-TN binary sequence spaces}
\noindent
We now apply the MS algorithm to the sets {\SI} and {\SII} which represent the full binary sequence spaces for the A1-N1 and A2-TN systems, respectively. Both sets include $2^{10}=1,024$ sequences. Using MS simulations, we determined the thermodynamic behavior at the temperatures $\TlowI$ and $\TlowII$ for the A1-N1 and A2-TN systems, respectively (see Table~1). In particular, this allow us to find the stabilities of folds IA and IB ($\PIA$ and $\PIA$) for all sequences in {\SI} and the stabilities of folds IIA and IIB ($\PIIA$ and $\PIIB$) for all sequences in {\SII}.

Having calculated these fold stabilities, we are in a position to determine if there are pathways in sequence space that lead to IA-IB or IIA-IIB fold changes without passing through an unstable intermediate sequence. To this end, we construct graphs in which each stable sequence is represented by a node and any two nodes are connected if their sequences differ at only one amino acid position. To determine if a sequence is stable we used the criterion $P_\mathrm{tot}>\Pcut$, where  $\Ptot = \PIA + \PIB$ and $\PIIA + \PIIB$ for A1-N1 and A2-AN1, respectively. The precise network obtained depends, of course, on the cut-off value of $\Pcut$. In general, a higher $\Pcut$ means a selection of more stable pathways. Fig.~5 shows the networks obtained with $\Pcut=0.50$. The number of viable mutational pathways that connect the start and end points are 516,972 between A1 and N1 and  57,912 between A2 and AN1. Because there are $10! =3,628,800$ possible mutational pathways between start and end points in both cases, these pathways represent 14.2 \% and 1.6 \% of all possible pathways in A1-N1 and A2-AN1, respectively. Hence, A1 and N1 are rather highly connected at this stability threshold. For $\Pcut=0.60$, the numbers are 104,640 pathways (2.9\%) for A1-N1 and 22,512 (0.6\%) pathways for A2-AN1. We find that there are no possible A1-to-N1 or A2-to-AN1 pathways when $\Pcut\ge0.74$ and $\Pcut\ge 0.66$, respectively. \\

@@

\section{Discussion}
\noindent
We have tested a generalized-ensemble algorithm for coarse-grained simulations of biomolecules that works by making the sequence a dynamic parameter, thereby leading to a random walk in sequence space. This approach was evaluated for protein folding simulations by comparing with ST. The results indicate two main benefits of the MS approach. On the one hand, it provides a convenient way to sample the respective canonical distributions for a large set of sequences in a single run and, on the other, it also enhances the sampling of conformational space.

The conformational sampling efficiency of the MS algorithm can be assessed from the comparison with ST in Figure~3. Although there is no single ``fair" way to compare MS and ST, we chose as a measure of efficiency the estimated statistical error on the total energy, $\sigE$, obtained with roughly the same computational cost per temperature and sequence. At the highest studied temperatures, which are well above the folding temperatures of both A1 and N1, the statistical errors $\sigE$ are basically the same for the two methods. This is not unexpected because conformational sampling of short polymers at high $T$ does not involve crossing any major energy barriers. As a result, successively sampled conformations for a given combination of sequence $\sbar$ and temperature $T$, are therefore likely mostly uncorrelated in both methods leading to similar $\sigE$ values.

At lower temperatures, we find that the MS simulations often yields significantly smaller $\sigE$ than ST simulations. It is notable that this acceleration in conformational sampling vis-{\'a}-vis ST is achieved despite that simulations are carried out at constant temperature. Hence, rather than promoting escape from local minima in the free energy landscape by visits to higher temperature, as is the case in  ST~\cite{Marinari1992,Lyubartsev1992} or replica-exchange,~\cite{Swendsen1986}  MS simulations escape local minima through visits to sequences for which the minima may not be as deep or are altogether absent. For the same reason, the performance of the MS algorithm will depend on the size and character of sequences included. In particular, the MS simulation likely hinge on the inclusion of at least some sequences with poor stabilities such that at least partial unfolding of the chain is regularly triggered and promoting transitions to new conformational states.

The MS method should not be seen as a general method to speed up conformational sampling due to the complexity of performing sequence updates in most biomolecular models. However, our results indicate that for coarse-grained models that permit the sequence updates as a Metropolis step and when visits to higher $T$ is unwanted (or unnecessary), the MS method can be highly efficient. A notable feature of the algorithm is that it provides escape from local minima on the free energy landscape of proteins without resorting to a change in $T$. This feature could make it attractive as an alternative to exploring the equilibrium behavior of proteins and peptides in the context or ordered dense phases such as lipid bilayers, where escape from local minima are especially challenging even in CG simulations.~\cite{Bereau2015} Simulations of lipid bilayers at elevated $T$ must typically be avoided because it can lead to unwanted perturbations of the membrane structure, although it can be a viable approach to speed up conformational sampling in particular cases.~\cite{Ulmschneider2010}
